---
title: "chapter6"
output: html_document
date: "2023-12-11"
---

# Analysis of longitudinal data

## Analysis of the RATS data

In the first part, I will analyse the RATS data, by providing graphical summaries and using linear mixed model.

```{r}
#read the data

RATS <- read.csv("RATS.csv")

#summary of the rats data
summary(RATS)

#Access the package ggplot2
library(ggplot2)
RATS$Group <- factor(RATS$Group)
RATS$ID <- factor(RATS$ID)

# Draw the plot
ggplot(RATS, aes(x = Time, y = RATS, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATS$RATS), max(RATS$RATS)))


```

I started by loading the data. the dataset contains longitudinal information on rats' weights.

It seems that in group 2 and 3 the starting weight is much higher than in group1. The increase seems to be about the same.

Next, I will get the group means for different timepoints and draw a plot on that data.

```{r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0)
library(dplyr)
library(tidyr)

#stabdardize the data
RATSS <- RATS %>%
  group_by(Time) %>%
  mutate( stdRATS = (RATS - mean(RATS))/sd(RATS)) %>%
  ungroup()


#number of participants per group
RATSS$n <- ifelse(RATSS$Group == 1, 8, 4)

#calculate standard error
RATSS1 <- RATSS %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(RATS), se = sd(RATS)/sqrt(n) )%>%
  ungroup()

# Plot the mean profiles
library(ggplot2)
ggplot(RATSS1, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(RATS) +/- se(RATS)")
```

Group 1 has smaller standard errors as other groups. This plot does not give much more information than the previous, group 2 and 3 probably increase a little faster than group 1.

Checking data outliers

```{r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0)
RATS1 <- RATS %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(RATS) ) %>%
  ungroup()

RATSS1 <- RATSS %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(RATS), se = sd(RATS)/sqrt(n) )%>%
  ungroup()

ggplot(RATS1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")
```

Well, in group 2 there is one outlier, but does not seem to be too far away.

T test sould not be performed as there are 3 groups, but lets do the anova.

```{r}

# Add the baseline from the original data as a new variable to the summary data
RATS2 <- RATS1 %>%
  mutate(baseline = RATS[RATS$Time == 1,]$RATS)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATS2)

anova(fit)
```

The weights increased statistically significantly from the baseline, but were not significantly different between groups.

## Analysis of BPRS data

First, load the data and plot the ratings per weeks.

```{r}
#read the data

BPRS <- read.csv("BPRS.csv")

#summary of the BPRS data
summary(BPRS)

#Access the package ggplot2
library(ggplot2)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

# Draw the plot
ggplot(BPRS, aes(x = week, y = BPRS, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRS$BPRS), max(BPRS$BPRS)))



```

Rating seem to be decreasing in both groups. Variability also decreasing.

Next let's create the regression model.

```{r}
model1 <- lm(BPRS ~ treatment + week, data = BPRS) 
model1
```

It shows large covariance with time and smaller with treatment groups. This means ratings are strongly decreasing as time goes and treament group 2 have littel higher scores.

Next, The Random Intercept Model.

```{r}
# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(BPRS ~ week + treatment + (1 | subject), data = BPRS, REML = FALSE)

summary(BPRS_ref)
```

Model quality seems to be high. similarly fixed effects show the same correlations.

Next, Random Slope Model and comparing it to the intercept model.

```{r}
#random slope model
BPRS_ref1 <- lmer(BPRS ~ week + treatment + (week | subject), data = BPRS, REML = FALSE)
#comparison with anova
anova(BPRS_ref, BPRS_ref1)

summary(BPRS_ref1)
```

P-value is 0.026, meaning that slope model fits well with the intercept model.

Next, lets add the intercept of time and treatment group

```{r}

BPRS_ref2 <- lmer(BPRS ~ week + treatment + (week | subject) + week*treatment, data = BPRS, REML = FALSE)

anova(BPRS_ref1, BPRS_ref2)

summary(BPRS_ref2)
```

Fits ok but not significant.

Last, plotting with raw and fitted values.

```{r}

#raw values
ggplot(BPRS, aes(x = week, y = BPRS, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "scores") +
  theme(legend.position = "top")

Fitted <- fitted(BPRS_ref2)

BPRS$fitted <- Fitted

#with fitted values
ggplot(BPRS, aes(x = week, y = fitted, group = subject)) +
  geom_line(aes(linetype = subject)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "scores") +
  theme(legend.position = "top")

```

Fitted values produce clearer plot as the values are adjusted for time and treatment group. Scores are decreasing as time goes
